Abstractions for Multi-Sorted Substitutions

### 2023-09-17 paper/paper.tex
Update.

@@

  With \emph{intrinsic typing}, the definitions of syntax and typing
  are merged into a single definition of well-typed terms.
  Consequently, the definition of an operation on terms also acts
  as a type preservation lemma for that operation, as it has to
  map well-typed terms to well-typed terms.
  For example, the definition of substitution
  also acts as a proof that substitution preserves typing, and the
  definition of a small-step semantics also acts as a proof for subject
  reduction.
  This often has the benefit of more concise code, as shared structure
  between definitions and preservation lemmas does not need to be
  repeated.

  With \emph{extrinsic typing}, the definitions of syntax and typing
  are kept separately: the syntax is defined first, and then typing is
  defined as a relation. Consequently, operations on terms and
  their type preservation lemmas are also separate entities.
  However, hope is not lost for extrinsic typing to keep up with the
  conciseness of intrinsic typing: 
  While extrinsic typing cannot eliminate the shared structure between
  definitions and preservation lemmas, it can automated by reflection
  or generic programming techniques, leaving only the preservation lemmas
  to be proven by the programmer. While typing can be arbitrary complex,
  the definition and metatheory of untyped substitution behaves the same
  across languages.

  However, when working with library support, this is less of a benefit.
  Most operations, like substitution, have a generic definition and
  metatheory across languages, which can be derived automatically via
  reflection or generic programming techniques.
  Only by adding a typing-relation

  However, intrinsic typing also has three major drawbacks:
  \begin{enumerate}
  \item
    It is more difficult to judge whether a specification is correct,
    because there is no clear separation between specification and proofs.
    To understand the specification one has to additionally
    understand the proofs, which can introduce significant cognitive overhead.
    % For example specifying a small step semantics is directly intertwined
    % with the proof of subject reduction.
  \item
    It is less uniform for object languages with multiple variable
    sorts, e.g. type- and expression-variables. In such a scenario,
    types are usually indexed by kinds and expressions by kinds and types.
    This requires not only a type-in-type and expression-in-expression
    substitution, but also a type-in-expression substitution, which causes
    interactions between different types of substitutions introducing
    additional complexity.
    The duplication increases even more if one additionally introduces
    kind-variables. Then a total of 6 different substitutions and
    their interactions is required.
    Unifying these substitutions to reduce repetition is difficult
    with intrinsic typing due to the interdependence of definitions.
  \item
    It is less suitable for automation, because type preservation is baked
    into the syntax definition and type preservation can become arbitrary complex
    depending on the typing relation. This can prevent substitition
    definition and lemmas from being derived, where otherwise it would be possible.
  \end{enumerate}
  With extrinsic typing the major drawback is the duplication between
  definitions and type preservation lemmas, which we found more
  suitable to be eliminated by automation.

  To combat those drawbacks, we went in the direction of intrinsically
  scoped syntax in our search for a general, clean, and automated framework
  for metatheory.

@@

  In this paper we present a simplified version of the actual
  framework, focusing on the core concepts. We present all necessary
  definitions and lemmas, but omit some proofs for the sake of
  space. The omitted proofs can be found in Appendix~\ref{sec:proofs} and the
  supplementary material. The full framework can be found at TODO.

@@

  \section{Syntax}
  \label{sec:syntax}
  \subsection{Single-Sorted Syntax}
  \label{sec:syntax:single}
  The following shows a typical intrinsically-scoped syntax of System F:
  \ExampleCode\FUnsortedSyntax

@@

  \section{Related Work}
  \label{sec:related}
  Autosubst \cite{DBLP:conf/cpp/StarkSK19, DBLP:conf/itp/SchaferTS15}.
  Kits \cite{DBLP:journals/jar/BentonHKM12, unpublished:mcbride2005kits}.
  Kits linear \cite{DBLP:journals/corr/abs-2005-02247}.
  Universe \cite{DBLP:journals/pacmpl/AllaisA0MM18}.

>>>>

  Formalizing programming languages in proof assistants quickly gets
  repetitive. Almost every programming language supports variables
  with static binding, and hence requires numerous definitions and
  lemmas related to variable substitution.

  Additionally, repetition can also occur within a single formalization.
  This can be seen with polymorphic languages, where multiple sorts
  of variables are present.
  Consider for example System F, which supports both expression- and
  type-variables.
  With a naive approach, the whole substitution machinery needs to be
  duplicated three times! We need to substitute expression-variables in
  expressions, type-variables in types, but we also need to substitute
  type-variables in expressions.
  Even worse, having two substitutions acting on expressions, requires
  to also prove lemmas about their interactions.
  If we would additionally introduce kind-variables, we would end up
  with a total of six duplications of the substition machinery and
  corresponding interaction lemmas!

  Further repetition can be introduced by the choice of variable
  representation. For example, for de Bruijn indices, substitution is
  usually defined in terms of renamings to allow for structural
  induction.
  With a naive approach, this would again double the amount of
  substitution machinery, as all definitions and lemmas need to be
  made first for renamings and then again for substitutions.

  Even more repetition is introduced by having a typed language,
  where for each substitution (and renaming) operation, we need to prove
  that the operation preserves typing, again doubling the repetition.

  % In summary, to have enough substitution machinery to prove subject
  % reduction for a language with expression-, type-, and
  % kind-variables one can end up with a total of 24
  % repetitions for each definition related to substitution, not counting
  % the interaction lemmas between different substitutions that are
  % acting on the same syntactic categories!
  In summary, formalizing a language with expression-, type-, and
  kind-variables with a naive approach can end up with a total of 24
  repetitions for each definition related to substitution, not counting
  the interaction lemmas between different substitutions that are
  acting on the same syntactic categories!

  In our framework we address those repetitions as followed:
  \begin{itemize}
  \item
    Instead of requiring a separate substitution for each sort of variable,
    we define a multi-sorted substitution which replaces all variables
    of all sorts simultaneously and can be applied to terms of all syntactic categories.
    This not only unifies many different definitions, removing redundancy on the way,
    but also eliminates the need for interaction lemmas.
  \item
    We use de Bruijn indices, but unify renamings and typings using the Kit abstraction.
    We further developed a Kit-like abstraction to also unify
    compositions of renamings and substitutions.
  \end{itemize}

  A close examination of the formalizations reveals that variable
  substitution behaves uniformly across languages, in the sense that
  the behavior of substitution is completely determined by the positions
  of variable bindings and variable uses.

@@

  With \emph{intrinsic typing}, the definitions of syntax and typing
  are merged into a single definition of well-typed terms.
  Consequently, the definition of an operation on terms also acts
  as a proof of type preservation for that operation, as it has to
  map well-typed terms to well-typed terms.
  For example, the definition of substitution
  also acts as a proof that substitution preserves typing, and the
  definition of a small-step semantics also acts as a proof for subject
  reduction.
  This often has the benefit of more concise code, as shared structure
  between definitions and preservation lemmas does not need to be
  repeated.

  With \emph{extrinsic typing}, the definitions of syntax and typing
  are kept separately: the syntax is defined first, and then typing is
  defined as a relation. Consequently, operations on terms and
  their type preservation lemmas are also separate entities.
  This has the benefit that specification and proofs are kept strictly separate,
  allowing a human to judge whether a specification is correct without also
  having to understand the proofs.
  % To understand the specification of intrinsic typing one has to
  % additionally understand the proofs, which can introduce significant
  % cognitive overhead.

  For our substition framework, we choose to work with extrinsic
  typing for a different reason.

  % For substitution, this separation brings an opportunity for automation:
  % free from the arbitrary complexity of typing relations,
  % the definitions and lemmas of substitution for untyped terms
  % behaves uniformly across object languages, and can be derived
  % completely.

  % While extrinsic typing cannot eliminate the shared structure between
  % definitions and preservation lemmas, it can automated by reflection
  % or generic programming techniques, leaving only the preservation lemmas
  % to be proven by the programmer. While typing can be arbitrary complex,
  % the definition and metatheory of untyped substitution behaves the same
  % across languages.

  % However, when working with library support, this is less of a benefit.
  % Most operations, like substitution, have a generic definition and
  % metatheory across languages, which can be derived automatically via
  % reflection or generic programming techniques.
  % Only by adding a typing-relation

  % However, intrinsic typing also has three major drawbacks:
  % \begin{enumerate}
  % \item
  %   It is more difficult to judge whether a specification is correct,
  %   because there is no clear separation between specification and proofs.
  %   To understand the specification one has to additionally
  %   understand the proofs, which can introduce significant cognitive overhead.
  %   % For example specifying a small step semantics is directly intertwined
  %   % with the proof of subject reduction.
  % \item
  %   It is less uniform for object languages with multiple variable
  %   sorts, e.g. type- and expression-variables. In such a scenario,
  %   types are usually indexed by kinds and expressions by kinds and types.
  %   This requires not only a type-in-type and expression-in-expression
  %   substitution, but also a type-in-expression substitution, which causes
  %   interactions between different types of substitutions introducing
  %   additional complexity.
  %   The duplication increases even more if one additionally introduces
  %   kind-variables. Then a total of 6 different substitutions and
  %   their interactions is required.
  %   Unifying these substitutions to reduce repetition is difficult
  %   with intrinsic typing due to the interdependence of definitions.
  % \item
  %   It is less suitable for automation, because type preservation is baked
  %   into the syntax definition and type preservation can become arbitrary complex
  %   depending on the typing relation. This can prevent substitition
  %   definition and lemmas from being derived, where otherwise it would be possible.
  % \end{enumerate}
  % With extrinsic typing the major drawback is the duplication between
  % definitions and type preservation lemmas, which we found more
  % suitable to be eliminated by automation.

  % To combat those drawbacks, we went in the direction of intrinsically
  % scoped syntax in our search for a general, clean, and automated framework
  % for metatheory.

@@

  In this paper we present a simplified version of the actual
  framework, focusing on the core concepts. We present all necessary
  definitions and lemmas, but omit some proofs for the sake of
  space. The omitted proofs can be found in Appendix~\ref{sec:proofs}
  and in the supplementary material. The full framework can be found at
  TODO.

@@

  \section{Syntax}
  \label{sec:syntax}
  \subsection{Unsorted Syntax}
  \label{sec:syntax:single}
  The following shows a typical intrinsically-scoped syntax of System F:
  \ExampleCode\FUnsortedSyntax

@@

  \section{Related Work}
  \label{sec:related}

  The kit abstraction for unifying renamings and substitutions
  appeared first in an unpublished article by
  McBride\cite{unpublished:mcbride2005kits}, and then later in
  Benton et al.\cite{DBLP:journals/jar/BentonHKM12}. 
  Wood and Atkey\cite{DBLP:journals/corr/abs-2005-02247} propose an
  extension to kits that supports substructural types via resource
  vectors.
  In all three cases they are formulated for intrinsic typing.

  \emph{Autosubst}\cite{DBLP:conf/itp/SchaferTS15}
  is a Coq framework, which derives parallel substitution definitions and
  lemmas for languages from annotated Coq syntax definitions using
  extrinsic typing, extrinsic scoping, and de Bruijn indices.
  The framework is implemented in Coq's tactic language LTAC and comes
  with a decision procedure for all assumption-free, equational
  substitution-lemmas.
  The implementation of Autosubst deals with multiple variable sorts
  by generating multiple substitutions and corresponding interaction lemmas.
  Autosubst does not provide any machinery for type preservation and
  does not unify renaming and substitution.
  % Assumes fun-ext.

  \emph{Autosubst 2}\cite{DBLP:conf/cpp/StarkSK19} is a standalone
  code generator, which translates second-order HOAS specifications into
  mutual inductive term sorts. Compared to Autosubst 1, it features
  mutually recursive object languages, intrinsically scoping, and
  vectorized substitutions.
  They also define and prove lemmas for all four compositions.
  The generated syntax consists of multiple term types, indexed by
  different amounts of free variables as shown in Section~\ref{sec:syntax}.
  % Does not assume fun-ext.

  Allais et al.\cite{DBLP:conf/cpp/Allais0MM17} propose a powerful
  abstraction for denotational semantics and proving semantic fusion
  lemmas.
  In later work\cite{DBLP:journals/pacmpl/AllaisA0MM18}, they use
  generic programming to instanciate this abstraction for a class of
  object languages comparable to ours.
  They demonstrate how both renaming and substitutions can be
  described as semantics and how the four composition laws follow from
  their generic fusion laws.
  They do not show how to work with multiple variable sorts, how to
  abstract over the compositions, and how to abstract over typing in
  an extrinsically typed scenario.

  Autosubst \cite{DBLP:conf/cpp/StarkSK19, DBLP:conf/itp/SchaferTS15}.
  Kits \cite{DBLP:journals/jar/BentonHKM12, unpublished:mcbride2005kits}.
  Kits linear \cite{DBLP:journals/corr/abs-2005-02247}.
  Universe \cite{DBLP:journals/pacmpl/AllaisA0MM18}.

  Abstract binding trees
